An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-07-20 nips_2016.tex
tweaks to norm2log table
added O(1/n) error bound

Here we derive a new test which incorporates the variability in
minibatch statistics as {\em a natural part of the test}. Because of
this, the amount of data required for each test is fixed with high
probability and in expected value. We use a Barker
test function \cite{Barker65} rather than a Metropolis test, which
makes the test naturally error tolerant. Similar ideas were presented
in \cite{Bardenet15} section 6.3. But the asymptotic test statistic
CDF and the Barker function are different, which leads to fixed
errors for the approach in \cite{Bardenet15}. Here we show that the
difference between the distributions can be corrected with an
additive random variable. This leads to a test which is fast, and
whose error can be made arbitrarily small.

@@

Its worth discussing at this point what are the goals (typically) of
M-H sampling on very large datasets.  By the Bernstein-von Mises
theorem, the posterior distribution of the parameter $\theta$ for a
Bayesian inference task is asymptotically normal, and has variance
that scales inversely with the number of data samples $N$. This mode
is extremely sharp for large datasets, which may contain millions or
billions of samples. Simply sampling from this distribution is one
application, but an efficient proposal distribution
\cite{OptimalScaling01} has similar variance to the target
distribution and will diffuse to it extremely slowly from an random
initialization value. If there are any other strong modes, it is very
likely for the sampler to find one of them and become trapped in it
when run at the normal distribution temperature (T=1). A common
solution is to anneal the sampler, running first at high temperature
(scaling log likelihoods by 1/T) which flattens the likelihood
landscape.  This is turn reduces the variance of the log acceptance
probability and allows our acceptance test to be applied.

A second question concerns step size. Once we have fixed temperature,
our variance constraint implies that we have to trade-off proposal
step size $s$ and batch size $b$ ($b \propto p^2$). i.e. we can make
many small steps, or one large step, with a given batch of data. One
of the primary drivers of this work is our belief in the value of
small steps. For applications to neural networks or other models where
the posterior is multimodel, posterior inference is arguably a search
process. Covering the search space densely with small steps is much
more valuable than few sparse steps toward the nearest optimum.

@@

\begin{lemma}\label{lem:efficienty_cutting}
  For the model above, there
  exists a fixed (independent of $N$) constant $c$ such that with probability $ \geq c$ over the joint
  distribution of $(\theta, \theta', u)$, the
  tests from \cite{icml2014c1_bardenet14} and \cite{cutting_mh_2014} consume all $N$ samples. 
\end{lemma}

@@

\begin{proof}
  Start with the LHS of (\ref{detbal}), substitute for $\alpha(\theta,\theta')$, and then apply the
  identity $g(s) = \exp(s)g(-s)$:
  \begin{equation}\label{detbal1}
    \begin{split}
      p(\theta)q(\theta'\mid\theta)\alpha(\theta,\theta') =~ & p(\theta)q(\theta'\mid\theta)g(\Delta(\theta,\theta'))\\
      =~ & p(\theta)q(\theta'\mid\theta)\exp(\Delta(\theta,\theta'))g(-\Delta(\theta,\theta'))
    \end{split}
  \end{equation}
  Note then that $-\Delta(\theta,\theta') = \Delta(\theta',\theta)$ because $\Delta$ is the log of ratios of
  terms involving $\theta$ and $\theta'$ and exchanging the two inverts all these ratios, negating the
  logarithm. Continuing:
  \begin{equation}\label{detbal2}
    \begin{split}
    p(\theta)q(\theta'\mid\theta)\exp(\Delta(\theta,\theta'))g(-\Delta(\theta,\theta')) = ~ & p(\theta)q(\theta'\mid\theta)\exp(\Delta(\theta,\theta'))g(\Delta(\theta',\theta))\\
    = ~ & p(\theta)q(\theta'\mid\theta)\exp(\Delta(\theta,\theta'))\alpha(\theta',\theta)
    \end{split}
  \end{equation}
  Substituting now for $\Delta(\theta,\theta')$ from equation (\ref{eq:delta0}) gives
  \begin{equation*}\label{detbal3}
    p(\theta)q(\theta'\mid\theta)\frac{p(\theta')q(\theta\mid\theta')}{p(\theta)q(\theta'\mid\theta)}\alpha(\theta',\theta)
      = p(\theta')q(\theta\mid\theta')\alpha(\theta',\theta) = RHS
  \end{equation*}
\end{proof}

As a sanity check, choosing $g(s) = \exp(s) \wedge 1$ produces the
standard Metropolis acceptance test $\alpha(\theta,\theta') = g(\Delta(\theta,\theta')) = 
\frac{p(\theta')q(\theta \mid
  \theta')}{p(\theta)q(\theta' \mid \theta)}\wedge 1$. $g(s)$ satisfies
the condition $g(s) = \exp(s)g(-s)$, and therefore detailed
balance. In fact it is the optimal acceptance function in terms of
acceptance rate, since it accepts with probability 1 for $\Delta > 0$.

\subsection{Logistic Acceptance Function}
For our new MH test we use the logistic function:
$g(s)=(1+\exp(-s))^{-1}$. Straightforward arithmetic shows that it
satisfies the condition in Lemma~\ref{lem:detailed_balance}.  While it
is slightly less efficient than the Metropolis test when used on the
full dataset, we will see soon that its smoothness allows it to
naturally tolerate substantial variance in its input argument. This in
turn will lead to a much more efficient test on subsets of data.

@@

In the limit, since $\Delta^*$ is normally distributed about its mean $\Delta$, we
can write
\begin{equation}\label{eq:relationship}
\Delta^* = \Delta + X_{\rm norm}, \quad X_{\rm norm} \sim \mathcal{\bar{N}}(0, \sigma^2(\Delta^*)),
\end{equation}
where $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ denotes a distribution which is approximately
normal with variance $\sigma^2(\Delta^*)$. 
But to perform the test in equation (\ref{theexacttest}) we want $\Delta + X$ for
a logistic random $X$ (call it $X_{\rm log}$ from now on). 
Our approach is to decompose $X_{\rm log}$ as
\begin{equation}\label{eq:deconvolution}
X_{\rm log} = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm}$ has zero-mean, Normal distribution with
variance $\sigma^2$, and $X_{\rm corr}$ is a zero-mean ``correction''
term with density $C_{\sigma}(X)$.  The two variables are added
(i.e., their distributions convolve) to form $X_{\rm log}$. It will be
convenient to express this relationship in terms of the CDFs $N_{\sigma}(X_{\rm
  norm}) = \Phi(X_{\rm norm}/\sigma)$ and $S(X_{\rm log})$, where
$\Phi(.)$ is the Normal CDF with zero mean and unit variance and
$S(.)$ is the standard logistic function.

@@

@@

Figure~\ref{fig:gauss_mix_2} (to the left) shows a histogram of the final minibatch sizes used by
the adaptive subsampling method each iteration (on a log scale for readability).  Usually, the method
can make a decision with 200 samples. Occasionally, however, it must use all 10000 points, resulting
in lots of wasted computation.  For this particular experiment, that happened eight times. The mean of adaptive minibatch size is 520 which is roughly close to $\sqrt{N}$ (i.e. 100). This number validates the claims of Theorem~\ref{thm:efficiency}. In
contrast, our method keeps the minibatch size around 200. Figure~\ref{fig:gauss_mix_2} (in the middle) shows a histogram of the minibatch size, and the mean of our minibatch size is 247. 
The third plot in Figure~\ref{fig:gauss_mix_2} investigates the number of times ${\rm std}(\Delta')
\ge 1.2$ for only consuming one minibatch data. For six minibatch sizes (50, 100, 150, 200, 250, and 300), we ran MCMC with
our MH test five times and averaged the number of ``violations.'' As the size increases, the
number of violations decreases, as expected.

Figure~\ref{fig:gauss_mix_3} (to the left) shows a histogram of ${\rm std}(\Delta')$. Figure~\ref{fig:gauss_mix_3} (to the middle) shows the confidence interval of our estimated ${\rm std}(\Delta')$ at each iteration \footnote{In order to show the confidence interval clearly, we use moving smooth to smooth the confidence interval, and plot each value by interval of 50 iterations}. The blue line represents the standard deviation comes from the complete data, which can be treated as the ``true'' ${\rm std}(\Delta')$, the green line represents our estimation of ${\rm std}(\Delta')$ from the minibatch, and the two red dash lines represent the confidence interval of ${\rm std}(\Delta')$ by adding / subtracting one ${\rm std(std(}\Delta'))$. This figure indicates that our estimation of ${\rm std}(\Delta')$ from one minibatch data is accurate. Figure~\ref{fig:gauss_mix_3} (to the right) shows the value of $\rho/\sigma^3$ at each iteration. Recall Theorem~\ref{thm:efficiency}, the number of data instance required is $\frac{4 \ell^2 C^2 \rho^2 }{\sigma^6 \varepsilon^2}$. This figure reveals that $\rho/\sigma^3$ is roughly a constant, which means the data required for each iteration is a constant value which has nothing to do with the $N$.

@@

In this appendix, we describe the following topics, with an emphasis on clarity and understanding:

\begin{itemize}[noitemsep]
    \item More information on Section~\ref{ssec:gaussians}.
    \item More information on Section~\ref{ssec:logistic}.
    \item More information on Section~\ref{ssec:nets}.
\end{itemize}

%% \section{Proofs}\label{app:proofs}

>>>>

Here we derive a new test which incorporates the variability in
minibatch statistics as {\em a natural part of the test}. Because of
this, the amount of data required for each test is fixed with high
probability and in expected value. We use a Barker test function
\cite{Barker65} rather than a Metropolis test, which makes the test
naturally error tolerant. The idea of using a noise-tolerant test
using Barker's test function were suggested but not explored
empirically in \cite{Bardenet15} section 6.3. But the asymptotic test
statistic CDF and the Barker function are different, which leads to
fixed errors for the approach in \cite{Bardenet15}. Here we show that
the difference between the distributions can be corrected with an
additive random variable. This leads to a test which is fast, and
whose error can be made arbitrarily small.

@@

Its worth discussing at this point what are the goals (typically) of
M-H sampling on very large datasets.  By the Bernstein-von Mises
theorem, the posterior distribution of the parameter $\theta$ for a
Bayesian inference task is asymptotically normal, and has variance
that scales inversely with the number of data samples $N$. This mode
is extremely sharp for large datasets, which may contain millions or
billions of samples. Simply sampling from this distribution is one
application, but an efficient proposal distribution
\cite{OptimalScaling01} has similar variance to the target
distribution and will diffuse to it extremely slowly from an
initialization value which is (likely to be) many standard deviations
away. If there are any other strong modes, it is very likely for the
sampler to find one of them and become trapped in it when run at the
normal distribution temperature (T=1). A common solution is to anneal
the sampler, running first at high temperature (scaling log
likelihoods by 1/T) which flattens the likelihood landscape.  This is
turn reduces the variance of the log acceptance probability and allows
our acceptance test to be applied.

A second question concerns step size. Once we have fixed temperature,
our variance constraint implies that we have to trade-off proposal
step size $s$ and batch size $b$ ($b \propto p^2$). i.e. we can make
many small steps, or one large step, with a given batch of data. One
of the primary drivers of this work is our belief in the value of
small steps. For applications to neural networks or other models where
the posterior is multimodel, posterior inference is arguably a search
process. Covering the search space densely with small steps is much
more valuable than few sparse steps toward the nearest optimum. In
this mode, Metropolis-Hastings can be used in similar fashion to
Stochastic Gradient Descent. The goal in SGD is to make gradual
progress to a posterior mode with each step, taking small steps so
that the cumulative displacement has progressively lower variance. A
substantial part of the computational work of MCMC on massive datasets
will be similarly in reaching a stationary distribution, which really
means finding a deep posterior mode. Taking noisy small steps will
nevertheless make steady progress to a posterior mode since their bias
is in that direction. We demonstrate this behavior with our experiments.

@@

\begin{lemma}\label{lem:efficienty_cutting}
  For the model above, there
  exists a fixed (independent of $N$) constant $c$ such that with probability $ \geq c$ over the joint
  distribution of $(\theta, \theta', u)$, the
  tests from \cite{icml2014c1_bardenet14} and \cite{cutting_mh_2014} consume all $N$ samples. 
\end{lemma}
Proof is given in the appendix.

@@

This result is used in \cite{Barker65} to define the Barker acceptance test. 
As a sanity check, choosing $g(s) = \exp(s) \wedge 1$ produces the
standard Metropolis acceptance test $\alpha(\theta,\theta') = g(\Delta(\theta,\theta')) = 
\frac{p(\theta')q(\theta \mid
  \theta')}{p(\theta)q(\theta' \mid \theta)}\wedge 1$. $g(s)$ satisfies
the condition $g(s) = \exp(s)g(-s)$, and therefore detailed
balance. In fact it is the optimal acceptance function in terms of
acceptance rate, since it accepts with probability 1 for $\Delta > 0$.

\subsection{Barker (Logistic) Acceptance Function}
For our new MH test we use the Barker logistic function:
$g(s)=(1+\exp(-s))^{-1}$ \cite{Barker65}. Straightforward arithmetic shows that it
satisfies the condition in Lemma~\ref{lem:detailed_balance}.  While it
is slightly less efficient than the Metropolis test when used on the
full dataset, we will see soon that its smoothness allows it to
naturally tolerate substantial variance in its input argument. This in
turn will lead to a much more efficient test on subsets of data.

@@

In the limit, since $\Delta^*$ is normally distributed about its mean $\Delta$, we
can write
\begin{equation}\label{eq:relationship}
\Delta^* = \Delta + X_{\rm norm}, \quad X_{\rm norm} \sim \mathcal{\bar{N}}(0, \sigma^2(\Delta^*)),
\end{equation}
where $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ denotes a distribution which is approximately
normal with variance $\sigma^2(\Delta^*)$. 
But to perform the test in equation (\ref{theexacttest}) we want $\Delta + X$ for
a logistic random $X$ (call it $X_{\rm log}$ from now on). In \cite{Bardenet15} it was proposed to
use $\Delta^*$ in a Barker test anyway and tolerate the fixed error caused by this approximation.

Our approach is to instead decompose $X_{\rm log}$ as
\begin{equation}\label{eq:deconvolution}
X_{\rm log} = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm}$ has zero-mean, Normal distribution with
variance $\sigma^2$, and $X_{\rm corr}$ is a zero-mean ``correction''
variable with density $C_{\sigma}(X)$.  The two variables are added
(i.e., their distributions convolve) to form $X_{\rm log}$. It will be
convenient to express this relationship in terms of the CDFs $N_{\sigma}(X_{\rm
  norm}) = \Phi(X_{\rm norm}/\sigma)$ and $S(X_{\rm log})$, where
$\Phi(.)$ is the Normal CDF with zero mean and unit variance and
$S(.)$ is the standard logistic function.

@@

In fact we can do better than this approximation by using a more
precise limit distribution, under an additional assumption. Let $\mu_i$ denote the $i^{th}$ moment, and $b_i$
denote the $i^{th}$ absolute moment of $X$. Bentkus et al.'s work on Edgeworth
expansions \cite{Bentkus97} equation 2.2 gives:

\begin{lemma} Let $X_1,\ldots,X_n$ be
a set of zero-mean, independent, identically-distributed random variables with sample mean $\hat{X}$ and with $t$ defined as in Lemma 3. Then
\begin{equation}\label{eq:clt-bounds}
 \sup_x\left|{\rm Pr}(t<x) - G\left(x, \frac{\mu_3}{b_2^{3/2}}\right)\right| \leq \frac{c(\epsilon,b_2,b_3,b_4,b_{4+\epsilon})}{n}
\end{equation}
where
\begin{equation}
G_N(x,y) = \Phi(x) + \frac{y(2x^2+1)}{6\sqrt{n}}\Phi'(x)
\end{equation}
\end{lemma}
Here the limit $G_N(x,y)$ is a normal CDF plus a correction term that
decays as $n^{1/2}$. Importantly, since $x^2\phi(x) = \phi^{''}(x) + \phi(x)$ where $\phi(x)=\Phi'(x)$, the correction term can be rewritten giving:
\begin{equation}
G_N(x,y) = \Phi(x) + \frac{y}{6\sqrt{n}}(2\phi^{''}(x)+3\phi(x))
\end{equation}
From which we see that $G_N(x,y)$ is a linear combination of $\Phi(x)$, $\phi(x)$ and $\phi^{''}(x)$.

@@

Figure~\ref{fig:gauss_mix_2} (to the left) shows a histogram of the final minibatch sizes used by
the adaptive subsampling method each iteration (on a log scale for readability).  Usually, the method
can make a decision with 200 samples. Occasionally, however, it must use all 10000 points, resulting
in lots of wasted computation.  For this particular experiment, that happened eight times.
The third plot in Figure~\ref{fig:gauss_mix_2} investigates the number of times ${\rm std}(\Delta')
\ge 1.2$ for only consuming one minibatch data. For six minibatch sizes (50, 100, 150, 200, 250, and 300), we ran MCMC with
our MH test five times and averaged the number of ``violations.'' As the size increases, the
number of violations decreases, as expected.

Figure~\ref{fig:gauss_mix_3} (to the left) shows a histogram of ${\rm
  std}(\Delta')$. Figure~\ref{fig:gauss_mix_3} (to the middle) shows
the confidence interval of our estimated ${\rm std}(\Delta')$ at each
iteration \footnote{In order to show the confidence interval clearly,
  we use moving smooth to smooth the confidence interval, and plot
  each value by interval of 50 iterations}. The blue line represents
the standard deviation comes from the complete data, which can be
treated as the ``true'' ${\rm std}(\Delta')$, the green line
represents our estimation of ${\rm std}(\Delta')$ from the minibatch,
and the two red dash lines represent the confidence interval of ${\rm
  std}(\Delta')$ by adding / subtracting one ${\rm
  std(std(}\Delta'))$. This figure indicates that our estimation of
${\rm std}(\Delta')$ from one minibatch data is
accurate. Figure~\ref{fig:gauss_mix_3} (to the right) shows the value
of $\rho/\sigma^3$ at each iteration.

@@

In this appendix, we describe the following topics:

\begin{itemize}[noitemsep]
  \item Proofs ommitted from the main text. 
    \item More information on Section~\ref{ssec:gaussians}.
    \item More information on Section~\ref{ssec:logistic}.
    \item More information on Section~\ref{ssec:nets}.
\end{itemize}

\section{Proofs}\label{app:proofs}

{\bf lemma 1}
{\em  For the model of section 2.1, there
  exists a fixed (independent of $N$) constant $c$ such that with probability $ \geq c$ over the joint
  distribution of $(\theta, \theta', u)$, the
  tests from \cite{icml2014c1_bardenet14} and \cite{cutting_mh_2014} consume all $N$ samples. }
